The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-10-03 Paper.qmd
Paper: Added accepted changes from current paper draft. Updated plots, tables, and in-text calls of model values to fit with new scheme.
Analysis script: Many changes.

---
title: "Child Signals of Need in Utila, Honduras: An Exploratory Study"
execute: 
  echo: false
  warning: false
format:
  html:
    toc: true
lightbox: true
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
---

```{r setup, include=FALSE}
# source("dictionaries.R")
source("analysisforpaper.R", local = knitr::knit_global()) # experiment with taking this away

@@

Despite the complexity of childhood and adolescence and their importance to adult outcomes, the behavioral strategies of children are often overlooked, especially when it comes to how they elicit support from parents and other caretakers [@gottlieb2000; @wells2003; @hirschfeld2002]. To gain support, children use signals to communicate their desires to parents. Infant signals of need, particularly infant cries, have been extensively studied in a range of populations [@barr1991; @fouts2005; @konner1974; @kruger2010; @kushnick2006; @meehan2017; @müller2021; @tronick1987], yet the full range of potential child signals of need, which likely extend beyond cries, has never been investigated in children of all ages. In particular, if and when children signal need with costly signals [@grafen1990; @spence1973; @zahavi1975a] has rarely been investigated [for one exception, see, @syme2023a].

In this paper, we use exploratory analysis on child signaling data from Utila Honduras to examine the hypothesis that children have evolved strategies to tailor their signals of need to 1) their valuation of a given resource, 2) the extent of conflicts of interest with caretakers, and 3) the presence of existing information about a child's need or potential to benefit from a resource so that they can credibly convey they are worth additional investment. Such strategies can lead to some combination of verbal requests for more support, facial sadness, crying, withdrawal, temper tantrums and self-harm, all of which may be effective in certain circumstances and are sometimes successful in eliciting more support from caretakers, despite a child's low status and formidability.

## Interdependence and investment within the family

@@

Care motivated by a high degree of fitness interdependence between offspring and at least one parent is seen in many animals [@clutton-brock1991; @webb1999]. However, humans are a uniquely interdependent species among the living primates [@tomasello2012] in that one’s fitness prospects depend greatly on a network of kin and other social allies [@apicella2019]. This need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities [@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from parents [@kramer2018], However, other adults like grandparents, aunts, uncles, and non-relatives have also been found to provide significant investment in children in many societies [@kramer2018; @sear2008], something which was likely favored due to a combination of indirect fitness benefits, reciprocity, or reputation management [@gurven2000] + (Schniter Preprint (permission needed). Children also often serve as alloparents, helping to raise their younger siblings or provide for them through food preparation or collection [@hrdy2009; @meehan2009; @sear2008], despite still requiring substantial amounts of investment themselves [@kramer2018a].

Although the fitness prospects of both children and familial caretakers are largely intertwined, evolutionary theory predicts that disagreements about the amount of investment from parents will be ubiquitous [@fouts2005; @godfray1995; @godfray1992; @trivers1974]. Most resources relevant to fitness outcomes are limited so that one unit of a resource invested in one area cannot be used in another. In response to this basic fact of life, humans, like other organisms, are expected to have evolved to efficiently allocate resources between somatic effort (i.e., growth and maintenance) and reproductive effort (i.e., mating and parenting effort) [@daly1983], something which often pits the fitness interests of parents against the fitness interests of offspring.

Conflicts of interest within families are largely structured by the likelihood an individual shares their genes with another. As a sexually reproducing species, humans have a 50% chance of passing down a given gene to each child. Each child, however, shares more genes through descent with themselves (100%) than with their full siblings (50% on average), half-siblings (25% on average), and stepsiblings (0%). Due to this asymmetry, children are often expected to attempt to elicit more investment from parents than is optimal from the perspective of the parent, while parents are expected to attempt to invest less than is optimal for each child in order to invest in their other children, whom parents are expected to value identically, all else being equal [@trivers1974].

Potentially cut this paragraph: While parent-offspring conflict and sibling competition are likely ubiquitous, the resources being contested will vary across human groups. In relatively egalitarian and food constrained populations, conflict over resources is likely to be centered on food, which probably explains the negative impacts siblings have on child age-specific heights and weights among the Shuar of Ecuador [@hagen2009], although similar effects have also been found in industrialized populations [@lawson2008]. In societies with greater inequality, competition is more likely to be over wealth, as is the case with older brothers in pastoralist communities like the Gabbra and Arsi Oromo limiting the marriage prospects of laterborn sons [@gibson2011; @mace1996]. Compared to large-scale investments like bride price or inheritance, little is understood about smaller transfers of wealth to younger children despite the potential importance of such transfers to 1) facilitate the development of embodied capital outside of growth and fat reserves or 2) help children succeed in their peer groups.

Conflicts of interest between children and parents may also involve how much children invest in the household. As previously mentioned, each child is expected to prefer greater investment in oneself than in one’s siblings, while parents, who are equally related to all their children, should often be expected to favor a more even distribution [@trivers1974]. This can lead to conflict with older children who might prefer to spend their time in ways which are likely to result in future direct fitness benefits (e.g., finding romantic partners or forming strong bonds with friends) rather than taking care of their siblings or acquiring or preparing food [@lowe2003; @syme2023a]. Conflicts of interest over investment in the household also appear to extend to when a child starts having their own children [@kushnick2023; @emlen1995], with parents potentially benefiting from delaying the reproductive careers of certain offspring in order to gain more care for their other children [@flinn1989; @emlen1992] or speeding up a child’s departure from the household if the child’s presence limits how much caretakers can invest in other children they deem more worthy of investment [@kushnick2023].

## Signaling

In order to elicit more support, children often use signals. Signals are a key source of information for individuals in social species and are commonly used to indicate one's value as a mate, one’s formidability to competitors, or one’s worthiness of investment to parents and other social partners [@caro2016; @maynard-smith2004; @preuschoft2000; @setchell2005; @zahavi1975a]. Signals differ from cues in that signals were favored by natural selection due to their ability to alter the behavior or perceptions of others in ways that benefit the signaler, whereas cues are pieces of information organisms use to influence their behavior or physiological state that did not evolve for communicative effect [@maynardsmith1995]. Crying is likely a signal of need[^1], for instance, while bleeding would be a cue of need.

[^1]: We expect “signals of need” to actually be signals of resource valuation. However, we use “signal of need” here to distinguish signals that involve attempts at eliciting support through revealing private information about one’s vulnerabilities from the more commonly studied signals which reveal private information about one’s positive qualities (i.e., those related to partner choice and intrasexual competition).

When conflicts of interest over investment exist, there may be much to gain from deception (i.e., exaggerating or completely fabricating one’s quality or need) so that one can receive the benefits that come from being seen as having a particular quality without actually doing so [@zahavi1993]. Although the risks of being deceived can be substantial, signal receivers may still benefit from paying attention to signals so long as the information is credible enough to inform one’s behavior in ways that are adaptive on average [@bliegebird2001; @grafen1990]. These benefits coupled with the likelihood of strong selection against individuals who respond to fake signals results in the expectation that signals within species will often need to benefit both signaler and receiver to persist evolutionary [@hasson1994; @maynard-smith2004].

Credible signals[\[2\]](#_ftn2) can evolve and be maintained in multiple ways. Despite reasonable expectations that such circumstances will be rare [@zahavi1993], signals can be produced and trusted in instances when there are no conflicts of interest over what information is conveyed [@guilford1995; @maynardsmith1995]. They may also arise when signals cannot be faked (i.e., indexical signals) or in instances when there are greater net payoffs to signalers who have the quality being signaled to a greater extent than others [@maynardsmith1995]. Although the latter can arise through differential benefits [@penn2020; @tiokhin2016] or a combination of differential costs and benefits (Bergstrom preprint), more attention has been paid to the importance of differential costs alone, with costly signaling being a possible pathway to credible signaling whenever signals are prohibitively costly for those who might fake the signal but still worth making for those with the underlying trait [@maynard-smith2004].

Credible signals work by revealing private information about one’s qualities to others. In the context of child signals of need, a child has access to information about their internal state, which is unavailable to others, that allows them to estimate their valuation of additional investment. By signaling in a credible way, the child reveals their resource valuation to potential caretakers who may be skeptical of the child’s claims due to the common conflicts of interest described above. With the information revealed by the signal, caretakers should generally be more likely to provide additional support due to their fitness interest in the child’s wellbeing. When this is the case, all parties are better off than they would be without the signal, with caretakers benefitting from the revelation of fitness-relevant information and the child benefitting from the increased likelihood of a beneficial response as long as the benefits outweigh the costs of the signal to child and parent.

However, even if a child was completely successful in conveying their needs to caretakers, it may not be enough to result in a beneficial response. This is because belief in one’s child’s need is not the only factor determining the form and amount of support given by parents. Instead, parents are expected to weigh the costs and benefits of support with other investment options [@daly1987; @daly1983]. Ancestrally, this likely often meant evaluating the viability of the child as well as how much a particular resource benefited one child relative to other children or future children [@clutton-brock1991; @daly1983; @dawkins1976; @hagen1999]. The former is thought to have resulted in children having to demonstrate their vigor early in life [@soltis2004], while the latter is expected to result in context-specific support [@beaulieu2008; @frankenhuis2022]. For example, one unit of a resource is often expected to be worth more to a newborn than an adolescent, favoring that investment flowing to the newborn all else being equal. However, if resources are in short supply, parents may benefit more from investing more in older children as they are closer to reaching adulthood and have higher reproductive value [@daly1983]. Additionally, a parent may also decide against investing in a child despite deeming investment in a child worthwhile if time is not right (e.g., if a mother preferred to breastfeed during rest rather than foraging) [@altmann1980]. When this is the case, outward conflicts may exaggerate the extent of a conflict of interest and involve parents teaching children ways to behave that benefit them both [@altmann1980].

@@

Crying is the most studied way by which children may signal need. Crying is a human universal [@gracanin2018] with few suspected non-signaling functions outside of the role vocal cries may play in the development of vocalization [@wells2003] and the functions tears have in terms of cleaning, protecting, and lubricating the eye [@murube2009; @sznycer]. Hunger, pain, anger, fear, and separation are all common causes of crying [@vingerhoets2013; @wells2003], although tear production also often happens with joyful events [@sznycer; @vingerhoets2013; @zickfeld].

Infant cries have been found to result in increased amounts of care, resources, and attention they receive and increased proximity to parents [@lin2012]. For example, Karo Batak infants who cried more were breastfed more, and the length of crying bouts predicted the time mothers breastfed [@kushnick2006]. Similarly, !Kung San mothers frequently respond to crying infants by providing comfort and breastfeeding access, with comforting behaviors, in general, being far more common than one’s meant to restrict infant behavior [@kruger2010]. Although this study found mothers were the most common responders, siblings, fathers, and other adults frequently stepped in so that 88% of crying bouts were responded to, with most of the cries not responded to lasting under ten seconds. Similar responsiveness has also been reported among the Aka [@hewlett1998], Mbendjele hunter-gatherers [@chaudhary2024], the Efe [@tronick1987], and the Ache, among whom young children have been described as “win\[ning\] most conflicts with parents simply by crying and whining” [@kaplan1987]. Studies examining the potential for long-term benefits are rare and difficult to achieve; however, during a time of high infant mortality in a sample of 13 Massai infants, those with easy temperaments were more likely to die (5/7) than those with difficult temperaments (1/6), something which is consistent with signals resulting in beneficial treatment [@devries1987].

Crying does not always succeed in eliciting benefits, however. Compared to the previously mentioned groups, the Ngandu [@hewlett1998] and the Hadza [@blurtonjones1993] have been found to be more likely to ignore the cries of infants and young children, and concerns over the risk of spoiling children by being too responsive seem to be reasonably common in some Western populations [@barnett2010; @burchinal2010; @chaudhary2024; @frankenhuis2022; @maute2018]. Crying can also lead to negative responses beyond being ignored [@barr1990; @soltis2004; @zeifman2017]. It has been found to increase parental distress, the likelihood of maternal depression, and the child’s risk of maltreatment [@frodi1985; @fujiwara2011; @howard2006; @lin2012; @soltis2004]. For example, in a sample of 3,259 parents of Dutch infants under 6 months old, 5.6% self-reported smothering, slapping, or shaking their babies in response to their crying, outcomes which were more likely among parents who were single, unemployed, or deemed their infant’s crying to be excessive [@reijneveld2004]. Excessive infant crying, i.e., colic, has also been found to be a risk factor for multiple negative outcomes [@zeevenhooven2018], including abuse, shaken baby syndrome [@kurth2011], and earlier cessation of breastfeeding [@howard2006].

Although widely accepted as a signal, there is no consensus on why emotional crying may work to elicit positive responses. One possibility is that the physiological costs of crying for newborns and infants [@rao1997; @thureen1998] may be prohibitively costly for faking to pay off. When this is the case, crying may serve as a credible of vigor [@furlow1997; @lummaa1998; @soltis2004; @wells2003], something which may be particularly important early in life due to increased risk of infanticide for children who have has poor survival prospects, due to either health conditions or resource constraints [@daly1984; @soltis2004].

Potentially combine and condense with paragraph below: As children age, their reproductive value increases due to reduced odds of mortality, leading to a decreased risk of infanticide (Hagen 2004 comment). For this reason, there is likely a transition period in which crying is primarily meant to provide information about how much a child values further investment. Many models have been put forward which attempt to explain human crying and begging in other species as credible signals of need due to the differential costs and benefits of crying and analogous vocalizations in other species as deterring the exaggeration of need [@godfray1991; @kilner1997; @parker2002]. Although differing in their details, a common theme is that the indirect fitness costs signalers incur from imposing costs on that parents and siblings through begging (be them energetic costs, survival costs, or reduced likelihood of future siblings) creates a floor to possible deception but falls short of preventing it [@daly1990; @wells2003].

For older children, the relatively insignificant energetic costs of crying may not be enough to result in credible signals in times of substantial conflicts of interest. For this reason, crying among older children may be more similar to adult crying, something which has been found to result in increased support and empathy but sometimes come at the cost of reputation and the potential for stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs may be a means by which crying can still be a credible signal [@sznycer] with moderate conflicts of interest if they dissuade potential fakers from making the signal.

### Withdrawal

Withdrawal has been suggested to be an underappreciated way those with low social power may credibly signal need [@syme2023a]. Due to their lower status, children have limited ways to change how others treat them. However, as children 1) are a major component of parents' inclusive fitness and 2) can help family members through allocare and other forms of labor, they provide benefits that are valuable and hard to replace. By withholding these benefits or putting the future of these benefits into question, a child simultaneously delays the benefits they get from this cooperation while imposing costs on those with a fitness interest in the child. As these costs are likely lower for those truly in need than those who are not, withdrawal may be a credible way to signal need if the costs to faking the signal make it unprofitable. In these cases, the end result is analogous to a labor strike, in which the signaler conveys to others their situation demands more support, and if they want to continue receiving benefits from the child, they need to upregulate their support of the child.

One of the key proximate mechanisms behind withdrawal is sadness or low mood [@scherer1994]\[\[3\]\](#\_ftn3). Sadness is a type of negative affect that is generally thought to arise from loss [@horwitz2017; @levenson1999; @smith1993], especially when one has little power to reverse or mitigate the loss on one’s own [@emanuel2023; @gunderson2021]. Sadness and low mood, like other forms of psychological pain, probably evolved to motivate victims of adversity to shift their attention to the causes of adversity to mitigate its negative fitness consequences and to learn to solve or avoid future such adverse events [@andrews2009; @nesse1990; @thornhill1989]\[\[MG12\]\](#\_msocom_12). Low mood, for example, is often thought to be adaptive in situations when activity is dangerous or unlikely to result in benefits, with anhedonia, pessimism, and fatigue all helping reduce behavior that may not be worth the costs or that may cause additional problems [@nesse2019; @nettle2009]. It may also encourage the reappraisal of one’s circumstances after new information comes to light that one’s current strategy is not working [@nesse2009].

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to indicate need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010]. However, facial expressions, along with tears, may not be enough to credibly signal need in times of significant conflicts of interest. In these instances, depression and suicidal behavior have been suggested to function as ways to credibly signal need due to the expectation that such states are costlier for those facing adversity than those who are not (for more detail on these approaches see: [@hagen2003; @rosenthal1993; @syme2016; @watson2002; @gaffney2022].

### Temper tantrums and self-harm

@@

As with the vocalizations that come with crying, temper tantrums might motivate caretakers to upregulate their support of the signaler due to the energetic costs of tantrums or the increased risk of predation stemming from the noisy display, both of which might result in credible signals if they are costly enough to not be worth faking [@lummaa1998]. In addition, tantrums also impose direct costs on caretakers who may be less able to do fitness enhancing tasks [@altmann1980] like finding or preparing food or taking care of other children.

Another possibility is that temper tantrums carry the risk or implication of offspring self-harm. For example, Utilian mothers reported in focus groups that some temper tantrums involved children banging their heads on things when they do not get what they want, something that has been reported in other populations [@belden2008; @burrows1953; @daniels2012; @dennis1940; @vandenakker2022]. Similarly, young chimpanzees (*Pan troglodytes*) have been observed to beat their heads on the ground when refused access to breast milk [@goodall1986], and American white pelican chicks (*Pelecanus erythrorhynchos*) have been found to throw themselves on the ground, flail violently, and bite at their own wings when begging for food [@schaller1964]. Such behaviors have been suggested to follow similar logic to depression and suicide attempts as they are likely relatively less costly to those who are truly in dire need compared to those who are not to the point faking is not worth the costs [@hagen2008]. In other words, the risk of injury from violent temper tantrums carries greater fitness costs to a child who is receiving enough support to develop high levels of embodied capital than it does to a child who is neglected to the point where their survival prospects are low. In this scenario, the risk of harm is likely not worth the potential benefits of increased support to the child receiving substantial investment, but it may be for the neglected child who may have few other options to compel others to invest in them.

@@

This study was designed to help establish the natural history of child signaling behavior in Utila, Honduras, by quantifying the frequencies of common child signals of need and various types of parental responses. From this, we sought to use signaling theory to examine 1) the causes of child signals of need, 2) how and why signal costs might vary, 3) how parents respond to indications of child need, 4) how signaling behavior changes with age and sex, 5) and if there were differences in child signaling and parental responses two groups who face different levels of social and economic adversity.

A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. This hypothisized causal, bi-directional relationship between investment and signaling makes it difficult to isolate the effects of signaling behavior on investment in children when using standard measures of child welfare (i.e., BMI, health metrics, and height for age). When experiments are not appropriate, longitudinal studies with carefully selected variables can help address this [@arjas2004]. However, our small sample size with signaling data and anthropometric data at two time periods limits our ability to do this in this study. For this reason, all analyses for this study were exploratory and meant to inform future studies of child signaling.

@@

Participants for this study were recruited by a variety of means. JH recruited early participants through convenience and snowball sampling. Participants were also recruited via Facebook posts and fliers which were sent home with children the Utila's Methodist school.

>>>>

---
title: "The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study"
execute:
  echo: false
  warning: false
format:
  html:
    toc: true
lightbox: true
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
    - kind: float
      reference-prefix: Table S
      key: stbl
      space-before-numbering: false
---

```{r setup, include=FALSE}
source("dictionaries.R", local = knitr::knit_global())
source("analysisforpaper.R", local = knitr::knit_global())

@@

Compared to chimpanzee mothers, mothers in hunter-gatherer societies have shorter interbirth intervals and produce offspring that require many more years of parental care [@Davison_2021; @Davison_2022]. Unlike chimpanzee offspring, human offspring in ancestral populations therefore often grew up with multiple older and/or younger siblings who were potential competitors for parental investment and yet who could also provide (or require) allocare [@xx, @yy].

Although the fitness prospects of children and their familial caretakers are largely intertwined, evolutionary theory predicts that disagreements about the amount of investment from parents and other close kin will be ubiquitous [@fouts2005; @godfray1995; @godfray1992; @trivers1974; @bossan2013] because each child shares more genes through descent with themselves (100%) than with their full siblings (50% on average), half-siblings (25% on average), and stepsiblings (0%). Due to this asymmetry, children are often expected to attempt to elicit more investment from parents than is optimal from the perspective of the parent, while parents are expected to attempt to invest less than is optimal for each child in order to invest in their other children, whom parents are expected to value identically, all else being equal [@trivers1974; @bossan2013].

While parent-offspring conflict and sibling competition are likely ubiquitous, the resources being contested will vary across human groups. In relatively egalitarian and food constrained populations, conflict over resources is likely to be centered on food, which probably explains the negative impacts siblings have on child age-specific heights and weights among the Shuar of Ecuador [@hagen2009], although similar effects have also been found in industrialized populations [@lawson2008]. In societies with greater inequality, competition is more likely to be over wealth, as is the case with older brothers in pastoralist communities like the Gabbra and Arsi Oromo limiting the marriage prospects of laterborn sons [@gibson2011; @mace1996]. Compared to large-scale investments like bride price or inheritance, little is understood about smaller transfers of wealth to younger children despite the potential importance of such transfers to 1) facilitate the development of embodied capital outside of growth and fat reserves or 2) help children succeed in their peer groups.

## Cooperation within the family

Humans are a uniquely interdependent species among the living primates [@tomasello2012] in that one’s fitness prospects depend greatly on a network of close kin and other social allies [@apicella2019]. This need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities [@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from mothers and fathers [@kramer2018]. However, grandparents, aunts, uncles, and non-relatives also provide significant investment in children in many societies [@kramer2018; @sear2008], which was likely favored due to a combination of inclusive fitness benefits, reciprocity, and reputation management [@gurven2000] + (Schniter Preprint (permission needed). Children also often serve as alloparents, helping to raise their younger siblings via direct care, food collection and preparation, and other family tasks [@hrdy2009; @meehan2009; @sear2008] despite still requiring substantial amounts of investment themselves [@kramer2018a].

Children’s potential to invest in their families can also cause conflicts with their parents. Older children might prefer to spend their time investing in romantic or other social relationships, or learning valuable skills, whereas parents might prefer them to care for their siblings, acquire or prepare food, or complete other household chores [@lowe2003; @syme2023a]. Conflicts of interest over investment in the family also appear to extend to when a child starts having their own children [@kushnick2023; @emlen1995], with parents potentially benefiting from delaying the reproductive careers of certain offspring in order to gain more allocare for their other children [@flinn1989; @emlen1992] or speeding up a child’s departure from the household so as to provide more parental investment to other children [@kushnick2023].

## Signals of needs and wants

In order to elicit more support, children often use signals of needs and wants. Across the tree of life, signals evolve to reveal agents’ private information to other agents when doing so benefits both the signaler and the receiver [@caro2016; @maynard-smith2004; @preuschoft2000; @setchell2005; @zahavi1975a]. Children often have needs and wants, where needs involve avoiding potential fitness costs, such as poor nutrition, injuries, or threats to status, and wants involve acquiring potential fitness benefits, such as highly nutritious foods and valuable artifacts. Needs and wants are often not apparent to caregivers, however, i.e., they are private information, and hence children must signal them to receive help. Signals differ from cues in that signals were favored by natural selection due to their ability to alter the behavior or perceptions of others in ways that benefit the signaler, on average, whereas cues are information organisms use to influence their behavior or physiological state that did not evolve for communicative effect [@maynardsmith1995]. Crying when injured is likely a signal of need[\[1\]](#_ftn1), for instance, whereas bleeding would be a cue of need.

When conflicts of interest over investment exist, there may be much to gain from deception (e.g., crying despite the lack of need) so that one could receive the benefits that come from being seen as needy [@zahavi1993]. Unchecked, deceptive signals could result in receivers ignoring all signals.

Credible signals can nevertheless evolve and be maintained in multiple ways. Signals can be trusted in instances when there are no conflicts of interest over what information is conveyed [@guilford1995; @maynardsmith1995]. They can also arise when signals cannot be faked (i.e., indexical signals). Credible signals can also evolve when the benefits of sending a signal outweigh the costs for individuals with the underlying quality or state, but otherwise the costs outweigh the benefits [@spence1973; @zahavi75; @bergstrom2024]. Such signals are often termed “costly”, although what is costly is deception, not necessarily signaling per se [for discussion, see @bergstrom2024].

In the context of child signals of need, a child has access to information about their internal state, which is unavailable to others, that allows them to estimate their valuation of additional investment. By signaling in a credible way, the child reveals their resource valuation to potential caretakers who may be skeptical of the child’s claims due to the common conflicts of interest described above. With the information revealed by the signal, caretakers should generally be more likely to provide additional support due to their fitness interest in the child’s wellbeing. When this is the case, all parties are better off than they would be without the signal, with caretakers benefitting from the revelation of fitness-relevant information and the child benefitting from the increased likelihood of a beneficial response.

However, even credible signals of needs or wants might not result in a beneficial response because belief in a child’s need is not the only factor determining the form and amount of support given by caregivers. Instead, caregivers are expected to weigh the costs and benefits of support against other options [@daly1987; @daly1983]. Ancestrally, this likely often meant evaluating the viability of the child as well as how much a particular resource benefited one child relative to other children, current or future[@clutton-brock1991; @daly1983; @dawkins1976; @hagen1999]. The former is thought to have resulted in children having to demonstrate their vigor early in life [@soltis2004], while the latter is expected to result in context-specific support [@beaulieu2008; @frankenhuis2022]. For example, one unit of a resource is often expected to be worth more to a newborn than an adolescent, favoring that investment flowing to the newborn all else being equal. However, if resources are in short supply, parents may benefit more from investing more in older children as they are closer to reaching adulthood and have higher reproductive value [@daly1983]. Additionally, a parent may also decide against investing in a child, despite deeming that investment worthwhile, if time is not right (e.g., if a mother preferred to breastfeed during rest rather than foraging) [@altmann1980]. When this is the case, outward conflicts may exaggerate the extent of a conflict of interest and involve parents teaching children ways to behave that benefit them both [@altmann1980]. Finally, caregivers often have better information about the costs and benefits of various items and actions, and therefore can make more beneficial decisions for a child than the child can herself (e.g., preventing a child from playing with matches).

Saved for now: Although differing in their details, a common theme is that the indirect fitness costs signalers incur from imposing costs on their that parents and siblings through begging (be them energetic costs, survival costs, or reduced likelihood of future siblings) creates a floor to possible deception but falls short of preventing it [@daly1990; @wells2003].

@@

@@

Crying is the most studied way children signal need. Crying is a human universal [@gracanin2018] with few suspected non-signaling functions outside of the role vocal cries might play in the development of vocalization [@wells2003] and the functions tears have in terms of cleaning, protecting, and lubricating the eye [@murube2009; @sznycer]. Hunger, pain, anger, fear, and separation are all common causes of crying [@vingerhoets2013; @wells2003], although tear production also often happens with joyful events [@sznycer; @vingerhoets2013; @zickfeld].

Infant cries increase amounts of care, resources, and attention infants receive, as well as increase their proximity to parents [@lin2012]. For example, Karo Batak infants who cried more were breastfed more, and the length of crying bouts predicted the time mothers breastfed [@kushnick2006]. Similarly, !Kung San mothers frequently respond to crying infants by providing comfort and breastfeeding access, with comforting behaviors being far more common than behaviors restricting infants [@kruger2010]. Although this study found mothers were the most common responders, siblings, fathers, and other adults frequently stepped in so that 88% of crying bouts were responded to, with the majority of the cries ignored lasting under ten seconds. Similar responsiveness has also been reported among the Aka [@hewlett1998], the Mbendjele [@chaudhary2024], the Efe [@tronick1987], and the Ache, among whom young children have been described as “win\[ning\] most conflicts with parents simply by crying and whining” [@kaplan1987]. Studies examining the potential for long-term benefits are rare and difficult to achieve; however, during a time of high infant mortality in a sample of 13 Massai infants, those with easy temperaments were more likely to die (5/7) than those with difficult temperaments (1/6), an outcome that is consistent with signals eliciting beneficial treatment [@devries1987].

Crying does not always succeed in eliciting benefits, however. Compared to the previously mentioned groups, the Ngandu [@hewlett1998] and the Hadza [@blurtonjones1993] have been found to be more likely to ignore the cries of infants and young children, and concerns that being too responsive will spoil children are reasonably common in some Western populations [@barnett2010; @burchinal2010; @chaudhary2024; @frankenhuis2022; @maute2018]. Crying can also lead to negative responses beyond being ignored [@barr1990; @soltis2004; @zeifman2017]. It has been found to increase parental distress, the likelihood of maternal depression, and the child’s risk of maltreatment [@frodi1985; @fujiwara2011; @howard2006; @lin2012; @soltis2004]. For example, in a sample of 3,259 parents of Dutch infants under 6 months old, 5.6% self-reported smothering, slapping, or shaking their babies in response to their crying, outcomes that were more likely among parents who were single, unemployed, or deemed their infant’s crying to be excessive [@reijneveld2004]. Excessive infant crying, i.e., colic, has also been found to be a risk factor for multiple negative outcomes [@zeevenhooven2018], including abuse, shaken baby syndrome [@kurth2011], and earlier cessation of breastfeeding [@howard2006].

Although widely accepted as a signal, there is no consensus on why emotional crying might elicit positive responses. As the costs and benefits of signaling and gaining investment change with age, it is likely the strategic logic of crying changes too. It could be vital for newborns to signal health and vigor to reduce the risk of neglect or infanticide experienced by children with poor survival prospects [@daly1984; @soltis2004]. As crying comes with physiological costs [@rao1997; @thureen1998], it could serve as a credible signal of vigor if these costs are too high for unhealthy infants to bear [@furlow1997; @lummaa1998; @soltis2004; @wells2003]. Alternatively, crying might impose costs on caregivers by, e.g., attracting predators or interfering with important activities, motivating caregivers to invest more to stop the crying [so-called blackmail or bargaining models; @xx, @yy]. More generally, contact calls between parents and offspring have evolved in numerous species, so cries might simply be a means for infants to elicit attention from caregivers [@mehr2021]. For older children, the relatively insignificant energetic costs of crying might not be enough to ensure it is a credible signal when there are conflicts of interest. For this reason, crying among older children might be more similar to adult crying, which can increase support and empathy but sometimes also reduces reputation and increases stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs could be a means by which crying can still be a credible signal of need if they dissuade potential fakers from making the signal [@sznycer] when there are moderate conflicts of interest.

### Sadness

Sadness is a type of negative affect that is generally thought to arise from loss [@horwitz2017; @levenson1999; @smith1993], especially when one has little power to reverse or mitigate the loss on one’s own [@emanuel2023; @gunderson2021]. Sadness and low mood, like other forms of psychological pain, probably evolved to motivate victims of adversity to shift their attention to the causes of adversity to mitigate its negative fitness consequences and to learn to solve or avoid future such adverse events [@andrews2009; @nesse1990; @thornhill1989]. Low mood, for example, is often thought to be adaptive in situations when activity is dangerous or unlikely to result in benefits, with anhedonia, pessimism, and fatigue all helping reduce behavior that may not be worth the costs or that may cause additional problems [@nesse2019; @nettle2009]. It may also encourage the reappraisal of one’s circumstances after new information comes to light that one’s current strategy is not working [@nesse2009].

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to indicate need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010].

### Temper tantrums

@@

As with the vocalizations that come with crying, temper tantrums might motivate caretakers to upregulate their support of the signaler due to the energetic costs of tantrums or the increased risk of predation stemming from the noisy display, both of which might result in credible signals if they are costly enough to deter faking [@lummaa1998]. In addition, tantrums also impose direct costs on caretakers who may be less able to complete important tasks [@altmann1980] like finding or preparing food or taking care of other children.

Another possibility is that temper tantrums carry the risk or implication of self-harm. For example, Utilian mothers reported in focus groups that some temper tantrums involved children banging their heads on things when they do not get what they want, behaviors that have been reported in other populations [@belden2008; @burrows1953; @daniels2012; @dennis1940; @vandenakker2022]. Similarly, young chimpanzees have been observed to beat their heads on the ground when refused access to breast milk [@goodall1986], and American white pelican chicks have been found to throw themselves on the ground, flail violently, and bite at their own wings when begging for food [@schaller1964]. Such behaviors s could credibly signal need as they are likely relatively less costly to those who are truly in dire need compared to those who are not, thus deterring deceptive signaling [@hagen2008].

@@

Although infant signals of need, particularly infant cries, have been extensively studied in many populations [@barr1991; @fouts2005; @konner1974; @kruger2010; @kushnick2006; @meehan2017; @müller2021; @tronick1987], the full range of potential child signals of need, which likely extend beyond cries, has never been investigated in children of all ages. In particular, how and when older children signal need has rarely been investigated [for one exception, see @syme2023a].

This study aimed to establish the natural history of child signaling in Utila, Honduras, by quantifying the frequencies of common child signals of need and various types of caregiver responses. From these, we used signaling theory to explore 1) the causes of child signals of need, 2) how and why signal costs might vary, 3) how parents respond to indications of child need, 4) how signaling behavior changes with age and sex, 5) and if there were differences in child signaling and parental responses in two subpopulations that face different levels of social and economic adversity. More specifically, we hypothesized that children have evolved strategies to tailor their signals of need to 1) their valuation of a given resource, 2) the presence of existing information about a child's needs and wants (i.e., potential to benefit from a resource) and 3) the extent of conflicts of interest with caretakers, which should influence signal cost. Such strategies can lead to some combination of verbal requests for more support, facial sadness, crying, withdrawal, temper tantrums and self-harm, all of which may be effective in certain circumstances and are sometimes successful in eliciting more support from caretakers, despite a child's low status and formidability.

A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. This hypothesized causal, bi-directional relationship between investment and signaling makes it difficult to isolate the effects of signaling behavior on investment in children, especially given the lack of prior studies on signaling by older children in low income settings.For this reason, all analyses of these unique data were exploratory and meant to lay the foundation for future studies of child signaling.

@@

Participants for this study were recruited through convenience sampling. JH recruited early participants with the help of primary informants and the use of snowball sampling. Participants were also recruited via Facebook posts and fliers which were sent home with children attending Utila's Methodist school.
